---
title: "Specification files and configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specification files and configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

groundplotqc checks tables against a specification: a data dictionary, its code lists
and, where they exist, a datasets table, translation tables and a lineage spec. This
article reads the fish survey example that comes with the package and pre-flights it.
It grows with later versions.

## Reading a specification

The fish survey example names its dictionary columns and types its own way, so
`gpq_column_map()` and `gpq_type_map()` say what they mean.

```{r}
library(groundplotqc)
example <- function(file) system.file("extdata", "examples", file, package = "groundplotqc")
fish <- gpq_read_spec(
  dictionary = example("fish_dictionary.csv"),
  code_lists = list(
    water_body = example("fish_water_body.csv"), gear = example("fish_gear.csv"),
    species = example("fish_species.csv")
  ),
  column_map = gpq_column_map(
    table = "table", attribute = "field", type = "kind", key_type = "key",
    reference = "parent", lookup = "codes", description = "notes"
  ),
  type_map = gpq_type_map(example("fish_types.csv")),
  sentinels = gpq_sentinels(
    numeric = c(missing = -99, not_applicable = -88),
    character = c(missing = "?", not_applicable = "~"),
    date = c(missing = "?", not_applicable = "~")
  )
)
names(fish)
fish$keys
```

Every coded attribute resolves to a code list:

```{r}
fish$code_list_map
```

## Pre-flight

`gpq_preflight()` checks the specification before any data is read. A check passes,
warns, stops, or doesn't run when its input wasn't given.

```{r}
results <- gpq_preflight(fish)
results[, c("rule_id", "outcome", "not_run_reason")]
```

A defect in the specification is a finding, not a failure to read. Here a code list
repeats a code, which warns:

```{r}
repeated <- gpq_read_spec(
  dictionary = example("fish_dictionary.csv"),
  code_lists = list(
    water_body = example("fish_water_body.csv"),
    gear = data.frame(gear = c("GN", "GN", "TN")),
    species = example("fish_species.csv")
  ),
  column_map = gpq_column_map(
    table = "table", attribute = "field", type = "kind", key_type = "key",
    reference = "parent", lookup = "codes", description = "notes"
  ),
  type_map = gpq_type_map(example("fish_types.csv"))
)
warned <- withCallingHandlers(
  gpq_preflight(repeated),
  gpq_preflight_warning = function(w) invokeRestart("muffleWarning")
)
subset(warned, outcome == "warn")
```

## MAGPlot's specification

The MAGPlot 2.0 layer compiles its specification files into the package with
`data-raw/build_magp_config.R`, which pre-flights them first and writes nothing if
pre-flight stops.
